---
title: "Loading & Setup"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Loading & Setup}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

This vignette covers the initial steps for working with stGradient: loading spatial transcriptomics data and launching the interactive polygon selector.

## Loading Xenium Data

The `load_xenium_seurat()` function provides a standardized way to load 10x Xenium Seurat objects with stable cell IDs and cancer/non-cancer annotations.
   
```{r load-data}
library(stGradient)

seurat_obj <- load_xenium_seurat(
  rds_path = "path/to/sample_seuratObject.rds",
  celltype_col = "singleR.predicted.id.brcaAtlas",
  cancer_labels = c("Cancer LumA SC", "Cancer LumB SC", "Cancer Her2 SC")
)
```

This function:

1. Reads a Seurat `.rds` file
2. Creates a stable cell-to-barcode mapping
3. Adds an `is_cancer` metadata column based on the cell type labels you supply

## Next Steps

Once you have your Seurat object loaded, proceed to the [Region Selection](region-selection.html) vignette to select ROI polygons and membership assignment.
