---
title: "Niche Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Niche Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

stGradient includes a complete niche analysis workflow for comparing spatial neighborhood composition between clinical groups (e.g., pCR vs non-pCR in neoadjuvant chemotherapy studies). The pipeline identifies cellular niches, quantifies their patient-level proportions, and tests for associations with outcome.

## Overview

The workflow proceeds in three steps:

1. **Build niches** — compute per-cell k-nearest-neighbor neighborhoods and cluster them
2. **Test associations** — Wilcoxon/permutation tests and logistic regression
3. **Visualize results** — composition plots, bar charts, and coefficient plots

## Running the Niche Pipeline

The `run_niche_pipeline()` function operates on a list of Seurat objects (one per patient/sample):

```{r niche-pipeline}
library(stGradient)

# Define clinical status table
pcr_status <- data.frame(
  patient_id = c("P1", "P2", "P3", "P4"),
  pCR = c(1L, 1L, 0L, 0L)
)

# Use a stable set of cell types across all samples
all_celltypes <- c("Cancer LumA SC", "T Cells CD4+", "T Cells CD8+",
                   "Macrophages", "Fibroblasts", "Endothelial")

# Intratumoral niches (cells inside polygons)
intra <- run_niche_pipeline(
  seurat_list = seurat_list,
  subset_value = TRUE,
  pcr_status = pcr_status,
  all_celltypes = all_celltypes,
  celltype_col = "singleR.predicted.id.brcaAtlas",
  polygon_col = "inside_polygon",
  k = 20,
  n_niches = 8
)

# Peritumoral niches (cells outside polygons)
peri <- run_niche_pipeline(
  seurat_list = seurat_list,
  subset_value = FALSE,
  pcr_status = pcr_status,
  all_celltypes = all_celltypes
)
```

### Key Parameters

- `subset_value` — `TRUE` for intratumoral, `FALSE` for peritumoral
- `k` — number of nearest neighbors for neighborhood computation
- `n_niches` — number of niche clusters to identify

### Output

`run_niche_pipeline()` returns a list with:
- `niche_prop_df` — patient × niche proportion data frame (with `patient_id` and `pCR` columns)
- `niche_composition` — cell type composition of each niche

## Statistical Testing

### Wilcoxon + Permutation Tests

```{r stat-tests}
intra_niches <- setdiff(colnames(intra$niche_prop_df), c("patient_id", "pCR"))
intra_X <- as.matrix(intra$niche_prop_df[, intra_niches, drop = FALSE])
intra_y <- intra$niche_prop_df$pCR

stat_intra <- run_niche_stat_tests(intra_X, intra_y, intra_niches, n_perm = 1000)
```

Returns a data frame with Wilcoxon p-values and permutation-based p-values for each niche.

### Univariate Logistic Regression

```{r logistic}
lr_intra <- run_niche_univariate_lr(intra_X, intra_y, intra_niches)
```

Fits one logistic regression per niche (proportion → pCR status) and returns coefficients, standard errors, and p-values.

## Visualization

### Stacked Composition

Per-patient niche proportion stacked bar charts, grouped by pCR status:

```{r stacked}
plot_niche_stacked_composition(intra$niche_prop_df)
```

### Grouped Bar Charts

Side-by-side mean proportions for pCR vs non-pCR with error bars:

```{r grouped}
plot_niche_grouped_bars(
  stat_intra, intra_X, intra_y, intra_niches,
  title = "Intratumoral Niches"
)
```

### Logistic Regression Coefficients

Forest plot of log-odds coefficients with confidence intervals:

```{r coefficients}
plot_niche_coefficients(lr_intra, title = "Intratumoral Niches")
```

## Comparing Compartments

A common workflow compares intratumoral and peritumoral niches:

```{r compare}
# Run both compartments
intra <- run_niche_pipeline(seurat_list, subset_value = TRUE, pcr_status = pcr_status,
                            all_celltypes = all_celltypes)
peri <- run_niche_pipeline(seurat_list, subset_value = FALSE, pcr_status = pcr_status,
                           all_celltypes = all_celltypes)

# Test both
stat_intra <- run_niche_stat_tests(intra_X, intra_y, intra_niches)
stat_peri <- run_niche_stat_tests(peri_X, peri_y, peri_niches)
```

## Next Steps

- [Visualization](visualization.html) — additional plotting utilities for QC and distance profiles
