---
title: "Visualization"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

stGradient provides a comprehensive set of plotting functions for visualizing spatial analysis results. This vignette covers all available visualizations organized by analysis stage.

## QC and Spatial Overview Plots

### Spatial QC Plot

Visualize cell types and ROI status on the spatial coordinates:

```{r qc-plot}
library(stGradient)

qc_out <- qc_plot_xenium(
  obj = seurat_obj,
  roi_col = "inside_polygon",
  celltype_col = "singleR.predicted.id.brcaAtlas",
  show = "both",
  zoom = TRUE,
  zoom_pad = 0.05,
  grey_outside_roi = TRUE
)

qc_out$p
```

The `show` parameter accepts `"roi"`, `"celltype"`, or `"both"` for side-by-side panels.

### ROI Composition Plots

Three complementary views of cell type composition relative to the region of interest:

```{r roi-composition}
# Cell type distribution inside ROI only
plot_roi_celltype_dist(
  df = qc_out$df,
  roi_col = "inside_polygon",
  celltype_col = "singleR.predicted.id.brcaAtlas",
  metric = "proportion",
  labels = "both"
)

# Stacked bar: inside vs outside composition
plot_roi_inout_celltype_stack(
  df = qc_out$df,
  roi_col = "inside_polygon",
  celltype_col = "singleR.predicted.id.brcaAtlas",
  metric = "proportion"
)

# Dodged bar: side-by-side comparison
plot_roi_inout_celltype_dodge(
  df = qc_out$df,
  roi_col = "inside_polygon",
  celltype_col = "singleR.predicted.id.brcaAtlas",
  metric = "count",
  top_n = 10,
  top_n_by = "inside"
)
```

## Distance Profile Plots

### Transcriptomic Distance Comparison

Compare separation metrics across inside modes:

```{r dist-comparison}
plot_distance_comparison(combined_profile)
```

### All Metrics Faceted

View all three separation metrics in a faceted layout:

```{r all-metrics}
plot_all_metrics(combined_profile)
```

### Cell Type Heatmap

Which cell types drive composition differences at each distance:

```{r heatmap}
plot_celltype_heatmap(celltype_contributions)
```

### Cell Type Proportions

Trajectory of top cell type proportions across distances:

```{r proportions}
plot_celltype_proportions(celltype_contributions, top_n = 5)
```

## DBSCAN Polygon Plots

Visualize detected tumor polygons overlaid on cell coordinates:

```{r dbscan-viz}
plot_dbscan_polygons(seurat_obj)
```

## PCF Plots

### PCF Heatmap

Pairwise cell type co-occurrence outside tumor regions:

```{r pcf-heatmap}
plot_pcf_outside_heatmap(pcf_summary, midpoint = 1)
```

### PCF Pair Envelopes

Monte Carlo envelope curves for statistical significance:

```{r pcf-envelopes}
plot_pcf_outside_pair_envelopes(
  seurat_obj = seurat_obj,
  polygons = auto_polygons,
  celltype_col = "singleR.predicted.id.brcaAtlas",
  nsim = 39,
  file = "pcf_envelopes.pdf"
)
```

### Single Pair PCF

Focused ggplot2 PCF curve for one cell-type pair:

```{r pcf-single}
plot_pcf_pair_gg(env_list[["TypeA_vs_TypeB"]])
```

## Gene-Distance Plots

### Volcano and Trend Plots

```{r gene-distance-plots}
dist_plots <- plot_distance_genes(dist_res, top_n = 12, sample_label = "Sample_01")
dist_plots$volcano
dist_plots$trends
```

### T-Cell Hotspot Overlay

```{r hotspot-plot}
plot_tcell_hotspots(seurat_obj, hs, celltype_col = "singleR.predicted.id.brcaAtlas")
```

## Niche Plots

### Stacked Composition

```{r niche-stacked}
plot_niche_stacked_composition(intra$niche_prop_df)
```

### Grouped Bars with Statistics

```{r niche-bars}
plot_niche_grouped_bars(stat_intra, intra_X, intra_y, intra_niches, title = "Intratumoral")
```

### Logistic Regression Coefficients

```{r niche-coef}
plot_niche_coefficients(lr_intra, title = "Intratumoral")
```

## Customization

All plotting functions return `ggplot2` objects that can be further customized:

```{r custom}
library(ggplot2)

p <- plot_all_metrics(combined_profile)
p + theme_bw() + labs(title = "Custom Title")
```

Use `patchwork` or `cowplot` to combine multiple plots:
   
```{r combine}
library(patchwork)

p1 <- plot_distance_comparison(combined_profile)
p2 <- plot_celltype_heatmap(celltype_contributions)

p1 / p2
```
